### 2020-12-17 paper/sex_selection_spacing-ver5.tex
Edit Conceptual Framework

% Economic conditions
The improvements in economic condition, especially the doubling of real wages,
are likely to affect fertility although the direction is ambiguous; empirically, 
higher female wages lowers fertility, whereas higher male wages increases fertility 
\citep{Hotz1997,schultz97}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as wages increase time becomes more expensive, and
people therefore work more and spend less time on non-wage earning activities, such as 
children or leisure.
The income effect captures that higher wages increase the available income, which
leads people to spend more time on time-intensive activities and less time working.
Since women spend substantially more time on childrearing than men, the substitution effect 
dominates for women but the income effect for men.

Higher female wages will also tend to shorten birth spacing if having children requires 
the mother to curtail her economic activities.
Shortening birth spacing allows parents to take advantage of economies of scale in 
childrearing---for example, looking after two children requires less than double the time 
needed for one child \citep{Vijverberg1982,Hotz1997}.

% Child outcomes
Parents care both about how many children they have and their outcomes, and with 
increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means a stronger incentive to invest in their children's
education, which increases the cost of children and lowers fertility. 
With lower expected mortality fewer births are required to reach a desired number 
of surviving children, which furthermore allows parents to invest more in each child.

An increased desired for better child outcomes will also tend to lengthen birth spacing,
provided it has positive effects on child outcomes.
The clearest example of a positive effect of longer spacing that very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

% Son preference 
Empirically, stronger son preference only marginally increases average fertility, but
differential stopping behavior still affects the sex composition and fertility across 
families and leads to shorter spacing and worse health outcomes for daughters  
when there are no sons in the household
\citep{repetto72,clark00,Whitworth2002,Basu2010,Jayachandran2011,Barcellos2014}.

Stronger son preference increases the use of sex selection and each sex-selective
abortion increases the expected birth spacing by 6--12 months.
After an abortion the uterus needs at least two menstrual cycles---approximately 
two months---to recover, or the subsequent risk of spontaneous abortion increases 
substantially \citep{zhou00b}.
Once conception can be attempted, the waiting time to conception is likely between
one and six months. 
Finally, sex-determination tests are reliable only from three months of gestation. 
Hence, it takes at least six months before the couple is at the same point in the next 
pregnancy as they were in the prior pregnancy when they decided to abort.

@@

% Factors/predictions
Although the explanations above are based on economic conditions, desired child outcomes,
son preference, and sex selection, none of these are available in the data, so I instead 
use female education and area of residence as the two principal explanatory variables in 
the empirical analyses.
Not only are these the closest available proxies, prior research shows that both higher 
female education and urbanization are associated with lower fertility and increased 
sex selection use, and, therefore, likely also changes birth spacing
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Furthermore, both variables are of interest given the tremendous changes in India; 
Figure \ref{fig:education_over_time} shows the substantial increase in female education
and the urban proportion of the population almost doubled from 18\% in 1960 to 35\% in 
2019 \citep{United-Nations2019}.

@@

Urban women are likely to have lower fertility and higher use of sex selection than women 
in rural areas.
The lower fertility is the result of higher cost of children and lower child mortality 
risk in urban compared to rural areas.%
\footnote{
However, holding parental education constant, children in Indian slums have worse
health outcomes than in rural areas \citep{Portner2018a}.
}
The higher use of sex selection comes partly from the lower fertility and partly from
higher income and better access to health care facilities in urban areas.

As women receive more education their potential wages increase, which should shorten 
birth spacing and lower fertility, but the low and declining female labor force 
participation in India suggests that families face little incentive to space children 
more closely together for economic reasons.
Even as women's education increased, their labor force participation stagnated or 
decreased and is now one of the lowest outside the Middle East 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018}.

The decline in female labor force participation appear to be driven by a combination
of a relatively small expansion in the sectors where women work and the income effect 
from rapidly increasing male wages and education dominating the substitution effect from 
higher female wages \citep{Klasen2015,Bhargava2018}.
The income effect can dominate because not only do men have a higher average level of 
education than women, men are also paid more than women for a given level of education, 
and the returns to education increases with education in India \citep{Agrawal2011}.
Hence, increasing male education will result in an larger increase in income than a 
similar increase in female education.

However, in a situation with increasing returns to male education and declining female labor 
force participation, we might still observe an association between higher female education 
and both lower fertility and increased use of sex selection, provided better-educated women 
are better at producing child human capital as suggested by \citet{Behrman1999}.
In this scenario, better-educated women are desired not for their ability to generate
income but for their ability to produce better-educated sons.
The higher return to male education provides an additional incentive to invest in sons'
education---with a better-educated mother one such investment---and also lowers desired 
fertility, makes sex selection more likely, and possibly increases birth spacing.

% Sanskritization
Finally, although the expansion in female education will likely bring in groups who 
initially have different norms, the process of ``Sanskritization'' implies that as 
lower-castes females gain access to education, they adopt higher-caste norms such as stronger son 
preference and a retraction from the formal labor market \citep{Srinivas1956}.
The declining female labor force participation suggests that this process still 
operates \citep{Abraham2013,Chatterjee2018}.

% Summary / hypotheses
In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Finally, "Sanskritization" implies that the changing composition of better-educated
women will not substantially change the use of sex selection.

>>>>

% Economic conditions
The improvements in economic condition, especially the doubling of real wages,
are likely to affect fertility, although the direction is ambiguous; empirically, 
higher female wages reduce fertility, whereas higher male wages increase fertility 
\citep{Hotz1997,schultz97}.
According to economic theory, the substitution and income effects' relative strengths 
determine the effect of higher wages on fertility.
The substitute effect captures that when wages increase, time becomes more expensive, and
people, therefore, work more and spend less time on non-wage earning activities, such as
children or leisure.
The income effect captures that higher wages increase the available income, which
leads people to spend more time on time-intensive activities and less time working.
Because women spend substantially more time on childrearing than men, the substitution
effect dominates for women while the income effect dominates for men.

Higher female wages may also shorten birth spacing if having children requires mothers to
curtail their economic activities.
Shortening birth spacing allows parents to take advantage of economies of scale in 
childrearing---for example, looking after two children requires less than double the time 
needed for one child \citep{Vijverberg1982,Hotz1997}.

% Child outcomes
With increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means a stronger incentive to invest in their children's
education, which increases the cost of having children, thereby lowering fertility. 
With lower expected mortality, parents need fewer births to reach the desired number of
surviving children, allowing parents to invest more in each child.

An increased desire for better child outcomes will also tend to lengthen birth spacing,
provided it has positive effects on child outcomes.
The clearest example of a positive effect of longer spacing that very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

% Son preference 
Stronger son preference has traditionally been associated with shorter spacing and worse
health outcomes for daughters when there no sons in the household
\citep{Whitworth2002,Jayachandran2011}.
Although son preference increases average fertility only marginally, differential stopping
behavior means that girls tend to end up in larger families, resulting in fewer resources
per child \citep{repetto72,clark00,Basu2010,Barcellos2014}.

However, with the advent of sex selection, stronger son preference may lead to longer
spacing because each sex-selective abortion adds 6–12 months to the length of the birth
interval.
After an abortion, the uterus needs at least two menstrual cycles---approximately 
two months---to recover, or the subsequent risk of spontaneous abortion increases 
substantially \citep{zhou00b}.
Once conception can be attempted, the waiting time to conception is likely between
one and six months. 
Finally, sex-determination tests are reliable only from three months of gestation. 
Hence, it takes at least six months before the couple is at the same point in the next 
pregnancy as they were in the prior pregnancy when they decided to abort.

@@

% Factors/predictions
I use female education and area of residence as the two principal explanatory variables in
the empirical analyses because those are the two closest proxies available in the data to
cover the explanations above. 
Prior research shows that both higher female education and urbanization are associated with 
lower fertility and increased sex selection use, and, therefore, likely also changes birth spacing
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Furthermore, both variables are of interest, given the tremendous changes in India. 
Figure \ref{fig:education_over_time} shows the substantial increase in female education,
and the urban proportion of the population almost doubled from 18\% in 1960 to 35\% in 
2019 \citep{United-Nations2019}.

@@

Urban women are likely to have lower fertility and higher use of sex selection than women 
in rural areas.
The lower fertility results from the higher cost of children and lower child mortality 
risk in urban compared to rural areas.%
\footnote{
However, holding parental education constant, children in Indian slums have worse
health outcomes than children in rural areas \citep{Portner2018a}.
}
The higher use of sex selection comes partly from the lower fertility and partly from
higher income and better access to health care facilities in urban areas.

Despite that higher potential wages from more education should shorten birth spacing and
lower fertility, the low and declining female labor force participation suggests little
economic incentive to space children closer together.
Even as women's education increased, their labor force participation stagnated or 
decreased and is now one of the lowest outside the Middle East 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018}.

The decline in female labor force participation appears to be driven by a combination
of a relatively small expansion in the sectors where women work and the income effect 
from rapidly increasing male wages and education dominating the substitution effect from 
higher female wages \citep{Klasen2015,Bhargava2018}.
The income effect can dominate because not only do men have a higher average level of 
education than women, men are also paid more than women for a given level of education, 
and the returns to education increases with education in India \citep{Agrawal2011}.
Hence, increasing male education will result in a larger increase in income than a 
similar increase in female education.

However, higher female education might still be associated with lower fertility and more 
sex selection, provided better-educated women are better at producing child human capital, 
as suggested by \citet{Behrman1999}. 
In a situation with increasing returns to male education, households may desire 
better-educated women because of their ability to produce better-educated sons, 
rather than because of their higher potential income. 
The higher return to male education provides an additional incentive to invest in sons' 
education---with a better-educated mother one such investment---and leads to lower desired 
fertility, makes sex selection more likely, and possibly increases birth spacing.

% Sanskritization
Finally, the process of "Sanskritization" implies that as lower-caste females gain access
to education, they adopt higher-caste norms such as stronger son preference and a
retraction from the formal labor market \citep{Srinivas1956}. 
The declining female labor force participation suggests that this process still 
operates \citep{Abraham2013,Chatterjee2018}.
Thus, although the expansion in female education will likely bring in groups who initially 
have different norms, we are unlikely to see substantial changes in how women with a given 
education level behave.

% Summary / hypotheses
In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated as their
household income increase the most, and because of their use of sex selection. 
Finally, "Sanskritization" implies that the changing composition of better-educated
women will not substantially change the use of sex selection.
